---
title: "example_code"
author: "Henrik Eckermann"
date: "8/12/2019"
output: html_document
---


```{r}
# includes ggplot and dplyr
library(tidyverse)
```



## Part II: Phyloseq data structure


 * **Load example data** of the gut microbiome profiling of 1006
     western adults from [microbiome R
     package](https://microbiome.github.io/tutorials/Data.html). The
     example data sets are already in the [phyloseq
     format](http://joey711.github.io/phyloseq/import-data). R
     provides many readily available tools to handle and analyze
     taxonomic profiling data that is provided in this format.
     
     
```{r}
library(microbiome)
data(atlas1006) 
print(atlas1006)

# rename the example data
pseq <- atlas1006
```


 * **Microbal abundances / OTU Table** retrieve the microbial (OTU)
   abundance table from the example data set (phyloseq object). This
   example data set has abundances at the genus level instead of OTU
   level. The taxonomic groupings in this 16S data set are only
   approximations to exact genus-level taxonomic classifications. Tip:
   microbiome::abundances or phyloseq::otu_table
   
   
```{r}
# full tables of abundances
otu.absolute <- abundances(pseq)
# or
otu.absolute <- otu_table(pseq)

# abundances results in a matrix object
abundances(pseq) %>% class()
otu_table(pseq) %>% class()

# print only 5 genera and first 10 samples
otu.absolute[1:5,1:10] 
```


 * **OTU tables** Retrieve the OTU table and investigate: 1) How many
   different samples and genus-level groups this phyloseq object has?
   2) What is the maximum abundance of Akkermansia in this data set?
   Tips: see data operations in the [microbiome
   tutorial](https://microbiome.github.io/tutorials/)
   
   
```{r}
# 1 
dim(otu.absolute) # 130 genera (rownumber) and 1151 samples (column number)

# 2
otu.absolute["Akkermansia", ] %>% max()
```


 * Draw a histogram of library sizes (total number of reads per
   sample). You can use the readcount function, or count the sum of
   reads per sample by using the colSums command applied on the otu
   table abundances. The example data set is from phylogenetic
   microarray, hence the numbers correspond to light intensity on the
   array rather than sequencing reads but the interpretation is
   analogous and we can ignore this detail for demonstration
   purposes. The key idea is that different samples may have different
   bacterial load. Check [Vandeputte et
   al. 2017](https://www.nature.com/articles/nature24460) for further
   discussion on the differences between absolute and relative
   quantification of microbial abundances.
   
   
```{r}
# option 1
pseq %>% readcount() %>% hist()
# option 2
otu.absolute %>% colSums() %>% hist()
```


 * **Taxonomy table** Retrieve the taxonomy table and print out the
   first few lines of it with the R command head(). Investigate how
   many different phylum-level groups this phyloseq object has? Tips:
   phyloseq::tax_table and unique.
   
   
```{r}
# print first few lines
pseq %>% 
  tax_table() %>% 
  head()

# how many different phylum-level groups?

# option 1
tax_table(pseq)[, 1] %>% 
    unique() %>% 
    length()

# option 2 
pseq %>% 
  tax_table() %>% 
  as.data.frame() %>% # we convert the matrix to a data frame for the next step
  distinct(Phylum) %>% # select only unique Phyla
  dim() # count
```


 * **Sample metadata** Retrieve sample metadata. How many unique
     subjects this data set has? What is the male/femal ratio? Draw a
     histogram of subject ages. Tips: microbiome::meta




## Part III: Data manipulation and exploration

 * **Subsetting** Pick a subset of the phyloseq object including only
     obese individuals who are over 40 years old. How many there are?
     Tips: subset_samples

 * **Transformations** The phyloseq data has absolute abundances. Use
     the microbiome::transform command to convert these into relative
     abundances. Compare Akkermansia abundance using the example data
     before and after the compositionality transformation (with a
     cross-plot, for instance). Finally, compare the results to CLR-
     transformed data (see e.g. [Gloor et
     al. 2017](https://www.frontiersin.org/articles/10.3389/fmicb.2017.02224/full))

 * **Visual exploration** Visualize the population distribution of
   abundances for the following taxa: Akkermansia, Dialister,
   Gemella. Do the same for CLR-transformed abundances. Tip:
   microbiome::plot_density / microbiome::transform
   
 * Experiment with other [phyloseq data manipulation
   tools](https://microbiome.github.io/tutorials/Preprocessing.html)